### 2018-04-16 paper.Rmd
comments on several sections

abstract: |
  The Child Language Data Exchange System (CHILDES) has played a critical role in research on child language development. Access to these data can be complex for novices and difficult to automate for advanced users, however. To address these issues, we introduce childes-db, a database-formatted mirror of CHILDES that improves data accessibility and usability by offering novel interfaces, including browsable web applications and an R application programming interface (API). Along with a versioned infrastructure that facilitates reproducibility of past analyses, these interfaces lower barriers to analyzing naturalistic parent-child language, allowing for a wider range of researchers in language and cognitive development to easily leverage CHILDES in their work.

@@

Children's language acquisition is a core puzzle in cognitive science. How are the representations underlying the productive use of language learned over children's first years? Because a language is comprised of many individual pieces of information -- the names of particular objects, the marker for the past tense, the position of determiners -- language development is fundamentally a question of learning from environmental input, which is the only source of information on these specifics. Further, children are often reluctant experimental participants; often their most revealing and most sophisticated uses of language emerge in naturalistic play. Thus, language acquisition research must be grounded in the ecological behavior of children, both their language input and their language production in context.

The foundational resource for studies of language acquisition is recordings and transcripts of children. Starting with Roger Brown's [-@brown1973] work on Adam, Eve, and Sarah, longitudinal recordings of children's talk (and the talk they hear from their caregivers) have formed the basis for this field. But creating individual transcripts is extremely hard work. Beyond gathering the data, even orthographic transcriptions (without phonetic detail) can take ten times the length of the original recording to transcribe [@macwhinney2000]. Thus, researchers' desire for data from multiple ages or multiple children can very quickly outstrip their resources.

Established in 1984 to address this issue, the Child Language Data Exchange System (CHILDES) aims to make transcripts and recording of children's language available for researchers via a free resource [@macwhinney2000]. CHILDES archives tens of thousands of transcripts of children’s speech, making it a critical resource for characterizing children’s early productive language use as well as their language environment. As the first major effort to consolidate and share transcripts of child language, CHILDES has been a pioneer in the move to curate and disseminate large-scale behavioral datasets publicly.

Since its inception, a tremendous corpus of research has made use of CHILDES data. Individual studies are too numerous to list, but classics include studies of morphological over-regularization [@marcus1992], distributional learning [@redington1998], word segmentation [@goldwater2009], the role of frequency in word learning [@goodman2008] and many others. Some studies analyze individual examples in depth [e.g., @snyder2007] while others work with the corpus as a whole [e.g., @meylan2017]. These are just a few of the thousands of studies that take CHILDES corpora -- transcripts, annotations, and associated media -- as starting points.

Yet there are some challenges working with CHILDES, both for students and for advanced users. The CHILDES ecosystem uses a special file format (CHAT), which is stored as plain text but includes structured annotations grouped into tiers (stored on separate lines) so that information about utterances can be stored with accompanying information such as the phonological or morphological structure of the utterance. These files are analyzed using a command-line tool (CLAN) with a host of specific arguments that allow users to count word frequencies, compute statistics (e.g., mean length of utterance, or MLU). While this system is flexible and powerful, mastering the CHAT codes and the CLAN tool can be daunting for a beginner or student. These technical barriers decrease the ease of casual exploration by a novice researcher or in a classroom exercise.

On the opposite end of the spectrum, for data-oriented researchers who are interested in doing large-scale analyses of CHILDES, the current tools are also not ideal. CLAN software is an excellent tool for interactive exploration, but -- as a free-standing application -- it can be tricky to build into a processing pipeline written in a modern analytic language such as Python or R. Thus, researchers who would like to ingest the entire corpus (or some large subset) into a computational analyses typically write their own parsers of the CHAT format to extract the subset of the data they would like to use [e.g., @redington1998;@yang2013;@meylan2017].

The practice of writing custom parsers is problematic for a number of reasons. Effort is wasted in rewriting the same parser again and again. In addition, this process introduces a major source of errors and inconsistencies in data handling due to difficulties dealing with the many special cases in the CHAT standard. Further, these parsing scripts are rarely shared, leading to much greater difficulty in reproducing the exact numerical results from papers using CHILDES [see e.g., @meylan2017 for an example]. In addition, the CHILDES corpus itself is a moving target; computational work using the entire corpus at one time point may include a different set of data than subsequent work (due to new contributions or revisions) and there is no simple way for researchers to document exactly what version of the corpus has been used short of creating a full mirror of the data. These factors together lead to a lack of computational reproducibility, a major problem that keeps researchers from verifying or building on published research [@donoho2010;@stodden2016].

In the current manuscript, we describe a system for extending the functionality of CHILDES to address these issues. Our system, childes-db, is a database-formatted mirror of CHILDES that allows programmatic access through an application programming interface (API). This infrastructure allows the creation of web applications for browsing and visualizing the data easily, facilitating classroom use of the dataset. Further, it can be accessed programmatically by advanced researchers, obviating the need to write one-off parsers of the CHAT format. The database is versioned for access to previous images, allowing computational reproducibility of particular analyses.

We begin by describing the architecture of childes-db and the web applications that we provide. We next describe the `childesr` API, which allows for programmatic access to the data. We then present several worked examples of specific uses of the system for research and teaching using both the web app and the API.

@@

# Using childes-db: Worked Examples

In this section we give a number of examples of how childes-db can be used in both research and teaching, using both the web apps and the R API.

@@

To investigate whether the frequency of color words is stationary over development, a user can navigate to the Frequency app, and enter a set of color words into the `Word` selector separated by a comma: here "blue, red, green." Because the question of interest is about the frequency of words in the input (rather than produced by children), the `Speaker` field can be set to reflect this choice. In this example we select "Mother." Because children learn most of their basic color words by the age of 5, the age range 1--5 years is a reasonable choice for `Ages to include`. The results of these selections are shown in Figure \ref{fig:color-shiny}. We can also create a hyperlink to store these set of choices so that we can share these results with others (or with ourselves in the future) by clicking on the `Bookmark` button in the bottom left corner.

From this figure, it seems likely that children hear "blue" more frequently early in development, but the trajectories of the "red" and "green" are less clear. We also do not have a good sense of the errors of these measurements, are limited to just a few colors at a time before the plot becomes too busy, and cannot combine frequencies across speakers. To perform this analysis in a more compelling and complete way, a user can use the `childesr` interface.

@@

We now join these two pieces of information together--how many times each speaker produced each color word, and how many total words they produced. We then group the data into 6-month age bins, and compute the proportion of tokens that comprise each color for each child in each 6-month bin. For comparability with the web app analysis, we convert the proportions to parts per million words.

@@

Gender differences in language production have been found in multiple studies, notably in @eriksson2012differences, a large-scale study that spanned ten languages. A researcher or student in child language acquisition may now investigate this difference with relative ease using `childesr`. By pulling data from the `transcript_by_speaker` table, a user has access to the the multitude of derived linguistic measures that are often used to evaluate a child's linguistic performance. In this worked example we walk through a sample analysis that probes the gender differences in language production.

First, we utilize the `childesr` function call `get_speaker_statistics` to pull data relating to the aforementioned derived measures for children and their transcripts. We exclusively select children and not their caregivers.

@@

This call pulls data from all collections and corpora, including those languages for which there are very sparse data. In order to make any substantial inferences from our analysis we begin by filterering languages for which there are a significant number of children. We also filter for children under the age of four years.

@@

Our `transcript_by_speaker` table contains multiple derived measures of language productivity and lexical diversity -- here we used MTLD [@mccarthy2005assessment], a measure that is robust to transcript length, as a way of evaluating children's linguistic productivity. We start by filtering those children for which a sex was defined, who speak a language in our subset, and are in the appropriate age range. We then average their MTLD score across transcripts after having grouped the data by child, age, sex, and language.

@@

<!-- AS: not sure whether I should have included this; I didn't want to overwhelm the reader-->

@@

@@

The `childesr` API provides instructors with an easy-to-learn method for giving students programmatic access to child language data. The aim is to reduce time spent getting data out of CHILDES and allow more course time for thinking about interesting questions that could be answered with NLP and child language data. Moreover, by writing our API in the R programming language, students will gain access to a general workflow can be integrated into an analysis pipeline that includes state of the art NLP tools.

For example, an instructor could create a programming assignment with the goal of reproducing the key findings in the case studies presented above -- color words or gender. Depending on the students' knowledge of R, the instructor could decide how much of the `childesr` starter code to provide before asking students to generate their own plots and write-ups. Moreover, the instructor could take advantage of recent advances in tools for teaching literate programming such as R Markdown [@xie2015dynamic], building in pre-written blocks of executable code that demonstrate key programming concepts for students and including incomplete or blank code blocks that students can use to write their own code.

@@

We have presented childes-db, a database formatted mirror of the CHILDES dataset. This database -- together with the `R` API and web apps -- facilitates the use of child language data. For teachers, students, and casual explorers, the web apps allow browsing and demonstration. For researchers interested in scripting more complex analyses, the API allows them to abstract away from the details of the CHAT format and easily create reproducible analyses of the data. We hope that these functionalities broaden the set of users who can easily interact with CHILDES data, leading to future insights into the process of language acquisition.

We have created childes-db to address a number of needs in our own teaching and research, but there are still a number of limitations that point the way to future improvements. For example, currently childes-db operates only on transcript data, without links to the underlying media files; these linkages facilitate deeper analyses of phonology, prosody, social interaction, and many other aspects of acquisition. Further, we have focused on including the most common and widely-used tiers of CHAT annotation into the database first, but our plan is eventually to include the full range of tiers. Finally, a wide range of further interactive analyses could easily be added to the current suite of web apps. We invite other researchers to join us in both suggesting and contributing new functionality as our system grows and adapts to researchers' needs.

>>>>

abstract: |
  The Child Language Data Exchange System (CHILDES) has played a critical role in research on child language development.
  Access to these data can be complex for novices and difficult to automate for advanced users, however.
  To address these issues, we introduce *childes-db*, a database-formatted mirror of CHILDES that improves data accessibility and usability by offering novel interfaces, including browsable web applications and an R application programming interface (API).
  Along with a versioned infrastructure that facilitates reproducibility of past analyses, these interfaces lower barriers to analyzing naturalistic parent-child language, allowing for a wider range of researchers in language and cognitive development to easily leverage CHILDES in their work.

@@

Children's language acquisition is a core puzzle in cognitive science.
How are the representations underlying the productive use of language learned over children's first years?
Because a language is comprised of many individual pieces of information -- the names of particular objects, the marker for the past tense, the position of determiners -- language development is fundamentally a question of learning from environmental input, which is the only source of information on these specifics. <!-- SM: innate knowledge? generalzation? -->
Further, children are often reluctant experimental participants; often their most revealing and most sophisticated uses of language emerge in naturalistic play. <!-- SM: we could motivate this differently-- it's not just that experiments are hard with young children; it's that many phenomena can only be seen in naturalistic data -->
Thus, language acquisition research must be grounded in the ecological behavior of children, both their language input and their language production in context.

The foundational resource for studies of language acquisition is recordings and transcripts of children. <!-- SM: this needs to be qualified--- Piaget, Berko, etc. experimental and diary studies predate naturalistic trancscripts -->
Starting with Roger Brown's [-@brown1973] work on Adam, Eve, and Sarah, longitudinal recordings of children's talk (and the talk they hear from their caregivers) have formed the basis for this field. <!-- SM: what field? child language? see above comment -->
But creating individual transcripts is extremely hard work.
Beyond gathering the data, even orthographic transcriptions (without phonetic detail) can take ten times the length of the original recording to transcribe [@macwhinney2000].<!--SM: qualify with "relatively simple" orthographic transcriptions; may need to define `phonetic` for general psych or (non-language) child dev audience -->
Thus, researchers' desire for data from multiple ages or multiple children can very quickly outstrip their resources.

Established in 1984 to address this issue, the Child Language Data Exchange System (CHILDES) aims to make transcripts and recording of children's language available for researchers via a free resource [@macwhinney2000].<!--SM: we might want to cite the whole set of CHILDES releases, starting with the earliest one in 1984 -->
CHILDES archives tens of thousands of transcripts and related media, making it a critical resource for characterizing children’s early productive language use as well as their language environment. <!-- SM: added a point about related media b/c Brian MacWhinney thinks this is a central point of CHILDES; tens of thousands of transcripts of children's speech +[across n languages] -->
As the first major effort to consolidate and share transcripts of child language, CHILDES has been a pioneer in the move to curate and disseminate large-scale behavioral datasets publicly.

Since its inception, a tremendous body of research has made use of CHILDES data. <!-- SM: switched corpus -> body to avoid confusion -->
Individual studies are too numerous to list, but classics include studies of morphological over-regularization [@marcus1992], distributional learning [@redington1998], word segmentation [@goldwater2009], the role of frequency in word learning [@goodman2008] and many others. <!-- SM: to convey its importance, we could say how many citations CHILDES main papers have received and how many citations papers that use CHILDES have received -->
Some studies analyze individual examples in depth [e.g., @snyder2007], other track multiple child-caregiver dyads [e.g., @meylan2017], and still others use the aggregate properties of child or caregiver speech. <!--SM: need a citation for using the aggregate speech --> 
These are just a few of the thousands of studies that take CHILDES corpora -- transcripts, annotations, and associated media -- as starting points.

Yet there are some challenges working with CHILDES, both for students and for advanced users.
The CHILDES ecosystem typically uses a special file format (CHAT), which is stored as plain text but includes structured annotations grouped into tiers (stored on separate lines) so that information about utterances can be stored with accompanying information such as the phonological, morphological, or syntactic structure of the utterance. <!-- added "typically" b/c BMW will note the availability of XML -->
These files are typically analyzed using a command-line tool (CLAN) with a host of specific arguments that allow users to count word frequencies, compute statistics (e.g., mean length of utterance, or MLU), and execture complex searches against the data.
While this system is flexible and powerful, mastering the CHAT codes and especially the CLAN tool can be daunting for a beginner or student.
These technical barriers decrease the ease of casual exploration by a novice researcher or in a classroom exercise.

On the opposite end of the spectrum, for data-oriented researchers who are interested in doing large-scale analyses of CHILDES, the current tools are also not ideal.
CLAN software is an excellent tool for interactive exploration, but -- as a free-standing application -- it can be tricky to build into a processing pipeline written in a data analysis platform such as the Python Pandas pacakage or R. <!--analytic -> data analysis platform -->
Thus, researchers who would like to ingest the entire corpus (or some large subset) into a computational analyses typically write their own parsers of the CHAT format to extract the subset of the data they would like to use [e.g., @redington1998;@yang2013;@meylan2017].

The practice of writing custom parsers is problematic for a number of reasons.
First, effort is wasted in implementing the same features again and again.
Second, this process introduces a major source of errors and inconsistencies in data handling due to difficulties dealing with the many special cases in the CHAT standard.
Third, these parsing scripts are rarely shared, leading to much greater difficulty in reproducing the exact numerical results from papers using CHILDES [see e.g., @meylan2017 for an example].
Fourth, the CHILDES corpus itself is a moving target; computational work using the entire corpus at one time point may include a different set of data than subsequent work (due to newly-contributed corpora or revised annotations of existing ones) and there is no simple way for researchers to document exactly what version of the corpus has been used short of creating a full mirror of the data.
These factors together lead to a lack of computational reproducibility, a major problem that keeps researchers from verifying or building on published research [@donoho2010;@stodden2016].

In the current manuscript, we describe a system for extending the functionality of CHILDES to address these issues.
Our system, childes-db, is a database-formatted mirror of CHILDES that allows programmatic access through an application programming interface (API). 
This infrastructure allows the creation of web applications for browsing and visualizing the data easily, facilitating classroom use of the dataset. 
Further, it can be accessed programmatically by advanced researchers, obviating the need to write one-off parsers of the CHAT format.
The database is versioned for access to previous releases, allowing computational reproducibility of particular analyses.

We begin by describing the architecture of childes-db and the web applications that we provide.
We next describe the `childesr` API, which allows for programmatic access to the data.
We then present several worked examples of specific uses of the system for research and teaching using both the web app and the API. 
<!-- a little too much "we" in this final paragraph; needs a final hopefil statement -->

@@

# Using childes-db: Worked Examples
<!-- SM: we should note somewhere that examples all use Tidyverse, so that R users who don't know what %>% don't trip -->
In this section we give a number of examples of how childes-db can be used in both research and teaching, using both the web apps and the R API.

@@

To investigate whether the frequency of color words is stationary over development, a user can navigate to the Frequency app, and enter a set of color words into the `Word` selector separated by a comma: here "blue, red, green."
Because the question of interest is about the frequency of words in the input (rather than produced by children), the `Speaker` field can be set to reflect this choice.
In this example we select "Mother."
Because children learn most of their basic color words by the age of 5, the age range 1--5 years is a reasonable choice for `Ages to include`.
The results of these selections are shown in Figure \ref{fig:color-shiny}.
We can also create a hyperlink to store these set of choices so that we can share these results with others (or with ourselves in the future) by clicking on the `Bookmark` button in the bottom left corner. <!--SM: do we need to update "Bookmark"" to "Share this analysis" ?-->

From this figure, it seems likely that children hear "blue" more frequently early in development, but the trajectories of the "red" and "green" are less clear.
We also do not have a good sense of the errors of these measurements, are limited to just a few colors at a time before the plot becomes too busy, and cannot combine frequencies across speakers. <!-- SM:"busy" seems a bit informal -->
To perform this analysis in a more compelling and complete way, a user can use the `childesr` interface.

@@

We now join these two pieces of information together--how many times each speaker produced each color word, and how many total words they produced.
We then group the data into 6-month age bins, and compute the proportion of tokens that comprise each color for each child in each 6-month bin.
For comparability with the web app analysis, these proportions are converted to parts per million words.

@@

Gender differences in language production have been found in multiple studies, notably in @eriksson2012differences, a large-scale study that spanned ten languages. <!-- SM what research questions specifically? many aspects of gender in language... probably better to note the range of gender-related effects in early language -->
A researcher or student in child language acquisition may now investigate this difference with relative ease using `childesr`. <!-- SM: unclear what "this difference" refers to -->
By pulling data from the `transcript_by_speaker` table, a user has access to the the multitude of derived linguistic measures that are often used to evaluate a child's linguistic performance. <!-- SM: linguistic performance -> grammatical development? -->
In this worked example we walk through a sample analysis that probes the gender differences in language production. <!-- SM: language production > gender differences in early grammatical development -->

First, we utilize the `childesr` function call `get_speaker_statistics` to pull data relating to the aforementioned derived measures for children and their transcripts.
Notee that we exclusively select the data of children, and exclude their caregivers.

@@

This call pulls data from all collections and corpora, including those languages for which there are very sparse data.
In order to make any substantial inferences from our analysis, we begin by filterering languages for which there are a significant number of children. 
We also filter for children under the age of four years.

@@

Our `transcript_by_speaker` table contains multiple derived measures of language productivity and lexical diversity -- here we used MTLD [@mccarthy2005assessment], a measure that is more robust to transcript length, as a way of evaluating children's linguistic productivity.
We start by filtering those children for which a sex was defined, who speak a language in our subset, and are in the appropriate age range.
We then average their MTLD score across transcripts after having grouped the data by child, age, sex, and language.

@@

<!-- AS: not sure whether I should have included this; I didn't want to overwhelm the reader -->
<!-- SM: need a few more details: what kind of model was it basically? loess, logistic growth curve, etc. -->

@@

<!-- SM: not terribly clear what the hazards were: sparsity and autocorrelation? -->

@@

The `childesr` API provides instructors with an easy-to-learn method for giving students programmatic access to child language data.
The aim is to reduce time spent getting data out of CHILDES and allow more course time for thinking about interesting questions that could be answered with NLP and child language data. <!-- SM: somewhat redundant with the above -->
Moreover, by writing our API in the R programming language, students will gain access to a general workflow can be integrated into an analysis pipeline that includes state of the art NLP tools. <!-- SM: sorry, what tools? there's very little NLP support in R b/c it's string handling is slow -->

For example, an instructor could create a programming assignment with the goal of reproducing the key findings in the case studies presented above -- color words or gender. 
epending on the students' knowledge of R, the instructor could decide how much of the `childesr` starter code to provide before asking students to generate their own plots and write-ups.
Moreover, the instructor could take advantage of recent advances in tools for teaching literate programming such as R Markdown [@xie2015dynamic], building in pre-written blocks of executable code that demonstrate key programming concepts for students and including incomplete or blank code blocks that students can use to write their own code. <!-- SM: cool, but scope creep? -->

@@

We have presented childes-db, a database formatted mirror of the CHILDES dataset.
This database -- together with the `R` API and web apps -- facilitates the use of child language data.
For teachers, students, and casual explorers, the web apps allow browsing and demonstration. For researchers interested in scripting more complex analyses, the API allows them to abstract away from the details of the CHAT format and easily create reproducible analyses of the data.
We hope that these functionalities broaden the set of users who can easily interact with CHILDES data, leading to future insights into the process of language acquisition. 
<!-- need a sentence on how it addresses complex use cases -->

We have created childes-db to address a number of needs in our own teaching and research, but there are still a number of limitations that point the way to future improvements.
For example, currently childes-db operates only on transcript data, without links to the underlying media files; these linkages facilitate deeper analyses of phonology, prosody, social interaction, and many other aspects of acquisition. <!-- SM: I thought we did have media links? -->
Further, we have focused on including the most common and widely-used tiers of CHAT annotation into the database first, but our plan is eventually to include the full range of tiers.
Finally, a wide range of further interactive analyses could easily be added to the current suite of web apps.
We invite other researchers to join us in both suggesting and contributing new functionality as our system grows and adapts to researchers' needs.
